---
title: "BIOSTAT 707 — Checkpoint 1: Set A cohort preparation and description"
author: "Bowen Zhang (bz147)"
date: now
engine: knitr
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: true
  cache: false
  error: false
  warning: false
  message: false
---

## Setup

This checkpoint preserves the reproducible long table and adds measurement-quality and temporal missingness analyses. It also constructs the initial wide table, Table 1, and descriptive outcome summaries. No prediction model is fitted.

```{r}
#| label: setup
here::i_am("checkpoint1.qmd")
library(data.table)
source(here::here("R", "load-set-a.R"))
source(here::here("R", "audit-missingness.R"))
source(here::here("R", "build-wide.R"))
source(here::here("R", "informative-missingness.R"))
library(ggplot2)
# Confirm that all available package libraries belong to the Pixi environment.
prefix <- normalizePath(Sys.getenv("CONDA_PREFIX"), mustWork = TRUE)
stopifnot(all(startsWith(normalizePath(.libPaths()), paste0(prefix, "/"))))
```

## Data locations and reproducibility

Run `pixi run --locked checkpoint1` from the repository root. The supplied
`pixi.toml` and `pixi.lock` specify the environment; raw data and generated CSVs
are excluded from Git. Set A must be supplied locally in `data/set-a/`.
Neither raw patient files nor outcomes are downloaded by this pipeline.

```{r}
raw_dir <- here::here("data", "set-a")
outcomes_path <- here::here("data", "Outcomes-a.txt")
output_dir <- here::here("output")
dir.create(output_dir, showWarnings = FALSE)
cat("Outcomes-a.txt present:", file.exists(outcomes_path), "\n")
```

Outcomes are validated locally. Their values are preserved, including the special Survival sentinel.

## What exists at prediction time?

Set A contains ICU admission descriptors and time-stamped observations from the
first 48 hours. These summaries would be available after that observation window,
not at admission. In-hospital death is the later outcome. Length of stay and
Survival also contain outcome information and are not predictors here.

The [Challenge documentation](https://physionet.org/content/challenge-2012/1.0.0/)
describes Event 1 as min(sensitivity, positive predictive value) and Event 2
as a Hosmer–Lemeshow-based calibration criterion. Neither is calculated at this
checkpoint because no prediction model has been fitted.

## Import Set A

All files must contain exactly `Time,Parameter,Value`. The loader reads these
columns as text, validates numeric values and elapsed times, and verifies one
internal RecordID per file against its filename. A malformed file stops the
render with a file-specific error. It does not silently skip records.

```{r}
imported <- load_set_a(raw_dir)
long <- imported$data
knitr::kable(data.frame(example_filename = basename(imported$files[c(1,1001,2001,4000)])))
```

## Long-table construction

Each source row is retained, including admission descriptors and the internal
RecordID row. `RecordID` is an identifier stored as text. `Time` is the original
elapsed-time string and `time_hours` is hours plus minutes divided by 60.
`Value_raw` retains the original value text (including its decimal formatting);
`Value` is numeric. Only numeric `-1` is changed to `NA` in `Value`.
The [PhysioNet Challenge 2012 documentation](https://physionet.org/content/challenge-2012/1.0.0/)
defines the elapsed-time convention and the missing/unknown sentinel.
No aggregation, deduplication, imputation, or outlier removal is performed.

```{r}
knitr::kable(head(long, 10))
```

## Data integrity checks

These checks execute before writing the CSV. Successful import means that all
`r length(imported$files)` files passed schema, time, numeric-value, and internal-RecordID checks.

```{r}
required <- c("RecordID", "Time", "time_hours", "Parameter", "Value_raw", "Value")
raw_numeric <- as.numeric(long$Value_raw)
sentinel <- raw_numeric == -1
stopifnot(length(imported$files) == 4000L,
          !anyDuplicated(basename(imported$files)),
          uniqueN(long$RecordID) == 4000L,
          identical(names(long), required),
          all(is.finite(long$time_hours)),
          all(is.finite(raw_numeric)),
          identical(is.na(long$Value), sentinel),
          identical(long$Value[!sentinel], raw_numeric[!sentinel]),
          long[Parameter == "RecordID", .N] == 4000L,
          all(long[Parameter == "RecordID", Value_raw == RecordID]))
validation <- data.table(
  check = c("Source files", "Unique RecordIDs", "Total rows", "Columns",
            "Filename/internal RecordID mismatches", "Schema/time/numeric parse failures",
            "Minimum elapsed hours", "Maximum elapsed hours", "Rows below 0 hours",
            "Rows above 48 hours", "Raw -1 values", "Cleaned NA values",
            "Raw values below -1", "Unique Parameters"),
  result = c(length(imported$files), uniqueN(long$RecordID), nrow(long), ncol(long),
             0, 0, min(long$time_hours), max(long$time_hours),
             sum(long$time_hours < 0), sum(long$time_hours > 48),
             sum(sentinel), sum(is.na(long$Value)), sum(raw_numeric < -1),
             uniqueN(long$Parameter)))
knitr::kable(validation)
parameter_counts <- long[, .(rows = .N), by = Parameter][order(Parameter)]
knitr::kable(parameter_counts, caption = "All unique Parameters and their row counts")
```

### Write and verify the output

`NA` is written explicitly in the cleaned numeric column. On import, specify
text types for the identifier and original fields to preserve their formatting.
The CSV is read back once, checking every column, row order, and numeric value
(with tolerance for CSV floating-point serialization).

```{r}
csv_path <- here::here("output", "set-a_long.csv")
fwrite(long, csv_path, na = "NA")
stopifnot(file.exists(csv_path))
roundtrip <- fread(csv_path, na.strings = "NA",
                   colClasses = list(character = c("RecordID", "Time", "Parameter", "Value_raw"),
                                     numeric = c("time_hours", "Value")))
stopifnot(nrow(roundtrip) == nrow(long), uniqueN(roundtrip$RecordID) == 4000L,
          identical(names(roundtrip), required))
for (column in c("RecordID", "Time", "Parameter", "Value_raw")) {
  stopifnot(identical(roundtrip[[column]], long[[column]]))
}
for (column in c("time_hours", "Value")) {
  stopifnot(isTRUE(all.equal(roundtrip[[column]], long[[column]], tolerance = 1e-12)))
}
fwrite(validation, here::here("output", "validation-summary.csv"))
fwrite(parameter_counts, here::here("output", "parameter-counts.csv"))
cat("CSV read-back passed:", nrow(roundtrip), "rows x", ncol(roundtrip), "columns.\n")
rm(roundtrip)
```

## Preliminary data-quality observations

### Measurement validity and data-quality decisions

The [PhysioNet variable definitions](https://physionet.org/content/challenge-2012/1.0.0/)
provide non-negative values, specific categories, and some explicit bounds.
SaO2 is a percentage, so its logical bounds are 0–100. These hard rules differ
from an analyst's plausibility screen. I use a separate `long_analysis` copy;
the six-column long CSV and both original value columns keep their Round 1 meaning.

```{r}
#| label: quality-audit
quality <- audit_measurements(long)
long_analysis <- quality$long
stopifnot(setequal(unique(long$Parameter),
                  c(time_series_parameters,"RecordID","Age","Gender","Height","ICUType")),
          length(time_series_parameters)==37L)
fwrite(quality$flags, here::here("output","data-quality-flags.csv"))
fwrite(quality$summary, here::here("output","data-quality-summary.csv"))
knitr::kable(quality$summary[flag_type=="hard",
  .(Parameter, validity_rule, violations, example_values, treatment)],
  caption="Documented hard validity rules (counts can overlap across rules)")
knitr::kable(long_analysis[hard_invalid == TRUE,
 .(observations=.N,patients=uniqueN(RecordID),values=paste(sort(unique(Value)),collapse=", ")),
 by=Parameter],caption="Distinct hard-invalid observations by variable")
```

There are `r sum(long_analysis$hard_invalid)` distinct hard-invalid observations.
Their `Value_analysis` is missing, but the rows, `Value_raw`, and `Value` remain.
I did not insert decimal points into out-of-range pH values or try to reinterpret negative temperatures. The
intended values cannot be recovered reliably from these records alone.
The `r sum(is.na(long$Value))` original sentinel values also stay missing.

### Soft plausibility concerns

The Height screen of 100–250 cm is a broad analyst-defined review range, **not**
an official validity range. Zero vital signs, weight, blood pressures, or PaO2
may reflect patient condition, recording practice, or monitoring problems.
They are flagged for review and still count as available measurements here.

```{r}
knitr::kable(quality$summary[flag_type=="soft",
 .(Parameter,validity_rule,violations,example_values)],caption="Soft concerns; values retained")
# Maxima are a rank-based inspection, not a threshold-based invalidity rule.
maxima <- quality$summary[flag_type=="soft_extreme_review",
 .(Parameter,rows_at_maximum=violations,maximum=example_values)]
knitr::kable(maxima,caption="Positive maxima of continuous variables: exploratory review only")
knitr::kable(head(quality$flags[flag_type %in% c("hard","soft"),
 .(RecordID,Time,Parameter,Value_raw,flag_type,reason)],8),caption="Example flags")
```

The maxima table helps identify possible unit or scale issues without declaring
that every maximum is wrong. The diagnostic CSV has one row per observation
and flag reason; a source observation may have more than one flag. `source_row`
identifies the corresponding row in the long table. No soft value is rescaled,
winsorized, deleted, or imputed.

## Missingness Analysis

### Definition and admission descriptors

A missing measurement often has **no row at all** in these skinny files.
Counting `is.na(Value)` would miss those absences. I count a variable as measured
when an admission has at least one nonmissing `Value_analysis` in the interval.
Every proportion uses all `r uniqueN(long$RecordID)` admissions, including those with no rows for
that variable. Hard-invalid values do not count; soft flags remain included.

Admission descriptors use observations at `00:00`. A later Weight does not
fill in a missing admission Weight. If multiple time-zero rows occur, at least
one valid value establishes availability; choosing a single descriptor value
is a later wide-table decision.

```{r}
ids <- sort(unique(long$RecordID))
descriptors <- admission_availability(long_analysis,ids)
fwrite(descriptors,here::here("output","admission-missingness.csv"))
knitr::kable(descriptors[,.(Parameter,n_measured,n_missing,missing_percent=round(100*p_missing,2))])
missingness <- measurement_availability(long_analysis,ids)
availability <- missingness$overall
temporal <- missingness$temporal
halves <- missingness$halves
fwrite(availability,here::here("output","availability-48h.csv"))
fwrite(temporal,here::here("output","availability-6h.csv"))
fwrite(halves,here::here("output","availability-24h-comparison.csv"))
# Soft flags do not change availability; hard exclusions can change it.
valid_rows <- long_analysis[Parameter %in% time_series_parameters,
 .(explicit_missing_rows=sum(is.na(Value_analysis)))]
cat("Explicit missing analysis rows among time-series rows:",valid_rows$explicit_missing_rows,"\n")
any_row_pairs <- uniqueN(long_analysis[Parameter %in% time_series_parameters &
 time_hours>=0 & time_hours<=48,.(RecordID,Parameter)])
absent_pairs <- 4000L*37L-any_row_pairs
total_missing_pairs <- sum(4000L-availability$n_measured)
stopifnot(absent_pairs<=total_missing_pairs)
knitr::kable(data.table(type=c("No row for admission-variable pair",
 "Rows exist, but none has a valid analysis value"),
 missing_pairs=c(absent_pairs,total_missing_pairs-absent_pairs)))

```

### Missingness over 48 hours

These are admission-level percentages, not percentages of recorded rows.
The 37 time-series variables include repeated Weight, but exclude RecordID,
Age, Gender, Height, and ICUType. The raw names TroponinI and TroponinT correspond
to TropI and TropT in the documentation.

```{r}
#| label: fig-overall-availability
#| fig-width: 8
#| fig-height: 10
#| fig-cap: "Patients with at least one valid measurement during 0–48 hours. Denominator: all admissions for every variable."
variable_order <- availability$Parameter
plot_overall <- ggplot(availability,aes(x=p_measured,y=factor(Parameter,levels=variable_order)))+
 geom_col(fill="#246987",width=.75)+
 scale_x_continuous(labels=scales::label_percent(),limits=c(0,1),expand=expansion(mult=c(0,.01)))+
 labs(x="Admissions with at least one valid measurement",y=NULL)+
 theme_minimal(base_size=11)+theme(panel.grid.major.y=element_blank())
ggsave(here::here("output","availability-48h.png"),plot_overall,width=8,height=10,dpi=160)
plot_overall
```

```{r}
knitr::kable(availability[,.(Parameter,n_measured,
 measured_percent=round(100*p_measured,2),missing_percent=round(100*p_missing,2))])
```

### Temporal missingness

Each heatmap cell is the proportion of `r length(ids)` admissions with at least one valid
measurement, regardless of how many measurements that patient had in the bin.
Bins are [0,6), [6,12), ..., [36,42), and [42,48]. Thus a reading at 06:00 starts
the second bin and 48:00 belongs to the last bin. Empty variable/bin cells
are explicitly included with zero availability.

```{r}
#| label: fig-temporal-availability
#| fig-width: 10
#| fig-height: 11
#| fig-cap: "Six-hour measurement availability. Rows are 37 time-series variables; columns are time windows; fill is the proportion of all admissions with a valid measurement."
bin_labels <- c("0–6","6–12","12–18","18–24","24–30","30–36","36–42","42–48")
plot_temporal <- ggplot(temporal,aes(x=factor(bin,levels=1:8,labels=bin_labels),
 y=factor(Parameter,levels=variable_order),fill=p_measured))+
 geom_tile(color="white",linewidth=.25)+
 scale_fill_viridis_c(limits=c(0,1),labels=scales::label_percent(),name="Admissions\nmeasured")+
 labs(x="Hours after ICU admission (48:00 included in final bin)",y=NULL)+
 theme_minimal(base_size=11)+theme(panel.grid=element_blank())
ggsave(here::here("output","availability-6h-heatmap.png"),plot_temporal,width=10,height=11,dpi=160)
plot_temporal
```

### First 24 versus second 24 hours

The first window is [0,24); the second is [24,48], including 48:00. These windows
do not overlap. Differences below are second minus first, in percentage points.

```{r}
half_display <- merge(availability[,.(Parameter,n_measured)],halves,by="Parameter")
setorder(half_display,n_measured,Parameter)
knitr::kable(half_display[,.(Parameter,first_percent=round(100*p_first,2),
 second_percent=round(100*p_second,2),difference_pp=round(difference_pp,2))])
largest_changes <- halves[order(-abs(difference_pp))][1:5]
knitr::kable(largest_changes[,.(Parameter,first_percent=round(100*p_first,2),
 second_percent=round(100*p_second,2),difference_pp=round(difference_pp,2))],
 caption="Five largest absolute changes between the two windows")
```

### Interpretation

Over 48 hours, availability ranges from `r round(100*min(availability$p_measured),1)`%
to `r round(100*max(availability$p_measured),1)`%. `r paste(tail(availability$Parameter,3),collapse=", ")`
are among the most routinely recorded variables, while
`r paste(head(availability$Parameter,3),collapse=", ")` are much less frequently
recorded. This is consistent with some tests being ordered selectively,
although these data alone do not establish why a test was ordered.

The largest absolute change between the first and second windows is for
`r largest_changes$Parameter[1]`: `r round(100*largest_changes$p_first[1],1)`%
to `r round(100*largest_changes$p_second[1],1)`%. Measurement availability
can change over the ICU window. I would keep first- and second-24-hour summaries
separate in the initial wide-table design so this timing is not lost.
This is a provisional data-summary choice, not a modeling decision.

Missingness here is largely the absence of measurements, rather than explicit
NA rows. Measurement patterns may reflect clinical workflow or patient
condition, so I would not automatically assume MCAR. This audit does not prove
MNAR or a causal explanation. These figures describe whether a measurement
exists, not how often it was repeated or whether it is clinically plausible.

## Outcome-file validation

The submitted pipeline reads the grader-provided `data/Outcomes-a.txt` and
contains no download step.

```{r}
outcomes <- fread(outcomes_path,colClasses=list(character="RecordID"),check.names=FALSE)
validate_outcomes(outcomes,ids)
outcome_validation <- data.table(rows=nrow(outcomes),unique_RecordIDs=uniqueN(outcomes$RecordID),
 matched_RecordIDs=sum(outcomes$RecordID %in% ids),
 in_hospital_deaths=sum(outcomes[["In-hospital_death"]]),
 death_proportion=mean(outcomes[["In-hospital_death"]]))
knitr::kable(outcome_validation)
fwrite(outcome_validation,here::here("output","outcome-validation.csv"))
```

No blanket `-1` conversion is applied to outcomes. In particular, Survival
records days to a recorded death; `-1` indicates a survivor under the Challenge
definition. The primary outcome is the supplied `In-hospital_death`, not a
reconstruction from Survival. Field-specific handling for descriptive summaries is explained below; the
original outcome values remain unchanged.

## Wide-table design

Round 2 showed substantial differences in measurement availability between the
first and second 24 hours. I therefore keep [0,24) and [24,48] separate while
creating one row per admission. Exactly 24:00 and 48:00 belong to the second
window; 00:00 belongs to the first. Every in-range observation has one window.

For each of the 37 variables in each window, count records the number of valid
observations. First and last retain some ordering information, min/max describe
the observed range, and mean gives the average recorded level. First/last use
time followed by source order for ties. These are summaries of the observed
measurements, not time-weighted averages. All valid tied observations contribute
to count/min/max/mean. This is an initial design that can be revised later,
not a claim that these are optimal model features.

Numerical summaries use `Value_analysis`: sentinel missing values and the `r sum(long_analysis$hard_invalid)`
hard-invalid observations are excluded; soft concerns remain. There is no
imputation, rescaling, or winsorization. A window with no valid measurements
has count 0 and NA for all five numerical summaries.

### Descriptor and timestamp audits

Admission descriptors use time zero only. A missing admission Weight is never
replaced by a later Weight. Both AdmissionWeight and repeated Weight summaries
are intentionally retained because Weight has both roles in the source data.
Identical descriptor duplicates would collapse to the first source row;
inconsistent duplicates stop construction for review rather than choosing a
value silently. Missing source rows would remain NA and appear in the audit.

```{r}
#| label: build-wide
built <- build_wide(long_analysis,ids,outcomes)
wide <- built$wide
descriptor_audit <- built$descriptor_audit
dup <- built$duplicates
fwrite(descriptor_audit,here::here("output","descriptor-extraction-audit.csv"))
fwrite(dup,here::here("output","duplicate-timestamp-audit.csv"))
descriptor_summary <- descriptor_audit[,.(expected_pairs=.N,
 missing_source_rows=sum(source_rows==0),duplicate_pairs=sum(source_rows>1),
 inconsistent_pairs=sum(distinct_values>1),minimum_rows=min(source_rows),
 maximum_rows=max(source_rows)),by=Parameter]
knitr::kable(descriptor_summary)
dup_summary <- data.table(duplicate_groups=nrow(dup),
 rows_in_duplicate_groups=sum(dup$rows),extra_rows=sum(dup$rows-1L),
 conflicting_valid_groups=sum(dup$valid_distinct>1))
knitr::kable(dup_summary)
```

The audit found `r nrow(dup)` duplicated patient/parameter/exact-Time groups;
`r sum(dup$valid_distinct>1)` contain conflicting valid values. Source order
makes first/last deterministic without averaging away those disagreements.
The raw observations stay intact in the long table.

## Wide-table construction and validation

Columns are RecordID, five admission descriptors, `r length(feature_columns())` time-series summaries,
and five outcome fields: **`r ncol(wide)` columns**. The summaries follow the fixed
37-variable order, first window then second window, and count/first/last/min/max/mean.
For example, `HR_h0_24_mean` and `HR_h24_48_count` identify the variable,
window, and statistic. `SAPS-I` is renamed `SAPS_I`, and `In-hospital_death` becomes `In_hospital_death`. Other outcome
names and values remain unchanged, including all raw outcome `-1` codes.

```{r}
wide_path <- here::here("output","set-a_wide.csv")
validate_wide(wide,ids,outcomes)
fwrite(wide,wide_path,na="NA")
wide_read <- fread(wide_path,na.strings="NA",colClasses=list(character="RecordID"))
validate_wide(wide_read,ids,outcomes)
stopifnot(isTRUE(all.equal(wide_read,wide,tolerance=1e-12,check.attributes=FALSE)))
knitr::kable(data.table(rows=nrow(wide_read),columns=ncol(wide_read),
 expected_columns=455,unique_RecordIDs=uniqueN(wide_read$RecordID),
 time_series_columns=length(feature_columns())))
```

Validation includes unique matching IDs, a one-to-one outcome join, unique
column names, no Inf/-Inf/NaN, non-negative integer counts, and the correspondence
between zero counts and five missing summaries. First/last use valid observations
only. Checks also confirm that mean lies between min and max.

`In_hospital_death` is the target. Length_of_stay and Survival contain
post-admission outcome information and must not become prediction features.
SAPS_I and SOFA are retained for cohort description/evaluation; they are not
automatically future predictors merely because they appear in this CSV.

## Wide-table missingness

This figure is calculated **from the read-back wide CSV**. Count is always
observed; count 0 means that first/last/min/max/mean are all missing for that
variable-window. If count is positive, all five are observed. Those five
summaries therefore share a missingness pattern, so plotting each separately
would repeat the same information. The fill here is missingness, whereas
Round 2's figures show availability.

```{r}
#| label: fig-wide-missingness
#| fig-width: 7
#| fig-height: 10
#| fig-cap: "Missingness derived from set-a_wide.csv: percentage of all admissions with count zero in each variable-window."
wide_missing <- wide_missingness(wide_read)
# Independently connect the new wide representation to the validated Round 2 counts.
check_half <- merge(wide_missing,halves,by="Parameter")
stopifnot(all(abs(check_half$p_missing-ifelse(check_half$window=="h0_24",
 1-check_half$p_first,1-check_half$p_second))<1e-12))
fwrite(wide_missing,here::here("output","wide-missingness.csv"))
plot_wide <- ggplot(wide_missing,aes(x=factor(window,levels=window_names,
 labels=c("0–24 h","24–48 h")),y=factor(Parameter,levels=variable_order),fill=p_missing))+
 geom_tile(color="white",linewidth=.35)+
 scale_fill_viridis_c(limits=c(0,1),labels=scales::label_percent(),name="Admissions\nmissing")+
 labs(x="Time window ([0,24) and [24,48])",y=NULL)+
 theme_minimal(base_size=11)+theme(panel.grid=element_blank())
ggsave(here::here("output","wide-missingness.png"),plot_wide,width=7,height=10,dpi=160)
plot_wide
```

### Admission-descriptor missingness from the wide table

```{r}
wide_descriptors <- c("Age","Gender","Height","ICUType","AdmissionWeight")
wide_descriptor_missing <- rbindlist(lapply(wide_descriptors,function(p)
 data.table(descriptor=p,missing_N=sum(is.na(wide_read[[p]])),
 missing_percent=100*mean(is.na(wide_read[[p]])))))
fwrite(wide_descriptor_missing,here::here("output","wide-descriptor-missingness.csv"))
knitr::kable(wide_descriptor_missing,digits=3)
```

## Table 1

Table 1 describes admission characteristics, stratified by the supplied
in-hospital death indicator. Continuous variables use median (Q1, Q3), and
categories use n (%) among nonmissing values within each column. Missing counts
are shown separately. There are no hypothesis-test p-values. Documented invalid
values are missing in analysis; broader plausibility flags remain because there
is not enough evidence to correct or remove them.

```{r}
#| label: table-one
#| echo: false
table_data <- as.data.frame(wide_read[,c(wide_descriptors,"In_hospital_death"),with=FALSE])
table_data$Gender <- factor(table_data$Gender,levels=c(0,1),labels=c("Female","Male"))
table_data$ICUType <- factor(table_data$ICUType,levels=1:4,
 labels=c("Coronary Care Unit","Cardiac Surgery Recovery Unit","Medical ICU","Surgical ICU"))
table_data$In_hospital_death <- factor(table_data$In_hospital_death,levels=c(0,1),labels=c("Survived","Died"))
table1 <- gtsummary::tbl_summary(table_data,by=In_hospital_death,
 type=list(c(Age,Height,AdmissionWeight)~"continuous",c(Gender,ICUType)~"categorical"),
 statistic=list(gtsummary::all_continuous()~"{median} ({p25}, {p75})",
                gtsummary::all_categorical()~"{n} ({p}%)"),
 digits=list(gtsummary::all_continuous()~1,gtsummary::all_categorical()~c(0,1)),
 label=list(Age~"Age (years)",Gender~"Gender",Height~"Height (cm)",
 ICUType~"ICU type",AdmissionWeight~"Admission weight (kg)"),
 missing="always",missing_text="Missing",missing_stat="{N_miss}") |>
 gtsummary::add_overall(last=FALSE) |>
 gtsummary::modify_caption("**Table 1. Admission characteristics by in-hospital outcome**") |>
 gtsummary::bold_labels()
fwrite(as.data.table(gtsummary::as_tibble(table1)),here::here("output","table1.csv"))
gtsummary::as_gt(table1)
```

Category labels follow the [official PhysioNet definitions](https://physionet.org/content/challenge-2012/1.0.0/).
The admission Weight column is independent of later repeated Weight summaries.

## Outcome summary

```{r}
cohort_totals <- data.table(total_admissions=nrow(wide_read),
 survived=sum(wide_read$In_hospital_death==0),died=sum(wide_read$In_hospital_death==1),
 mortality_percent=100*mean(wide_read$In_hospital_death==1))
stopifnot(cohort_totals$survived+cohort_totals$died==4000L)
knitr::kable(cohort_totals,digits=2)
severity <- outcome_description(wide_read)
fwrite(severity,here::here("output","outcome-descriptive-summary.csv"))
fwrite(cohort_totals,here::here("output","cohort-totals.csv"))
knitr::kable(severity[,.(field,nonmissing_N,missing_N,
 `Median (Q1, Q3)`=sprintf("%.1f (%.1f, %.1f)",median,q1,q3))])
```

For this table only, the documented missing/unknown code `-1` in SAPS_I, SOFA,
and Length_of_stay is treated as missing in separate copies. Their raw values
remain in the wide CSV. Length of stay describes the hospitalization rather
than a baseline characteristic. Survival is omitted from this numerical table:
its `-1` has special survivor semantics and is not an ordinary number of days.
Mortality comes directly from In_hospital_death.

## Is the missingness informative?

For each time-series variable, an admission is ever measured when it has at
least one valid `Value_analysis` during 0–48 hours, including both endpoints.
Sentinels and hard-invalid values do not count; soft flags remain included.
Each admission contributes once to either the measured or unmeasured group.
Death rates use the supplied in-hospital outcome. Differences below are signed
percentage points (measured minus unmeasured), sorted by absolute magnitude.
An unavailable rate is shown as NA when its group is empty.

```{r}
informative <- informative_missingness(long_analysis, ids, outcomes)
stopifnot(all(informative$n_measured + informative$n_not_measured == length(ids)),
          all(informative$deaths_measured + informative$deaths_not_measured ==
              sum(outcomes[["In-hospital_death"]])))
availability_check <- merge(informative, availability, by="Parameter")
stopifnot(all(availability_check$n_measured.x == availability_check$n_measured.y))
fwrite(informative, here::here("output", "informative-missingness.csv"), na="NA")
knitr::kable(informative[, .(Variable=Parameter, `Measured N`=n_measured,
 `Measured death %`=death_pct_measured, `Not measured N`=n_not_measured,
 `Not measured death %`=death_pct_not_measured,
 `Difference (pp)`=difference_pp)], digits=2)
```

Across the `r nrow(informative)` variables, mortality differs between measured
and unmeasured groups for some variables. Whether a measurement is taken may
reflect patient condition and clinical workflow, so I would not automatically
treat measurement absence as random. Small unmeasured groups can also produce
large, unstable differences. This descriptive comparison does not establish why
measurements are missing, establish causality, or prove a formal missingness
mechanism. The ordering is only a way to display the table; no hypothesis tests
or p-values are used.

## AI use disclosure

I used Codex/ChatGPT substantially for project and repository setup guidance,
R code drafting, data wrangling, missingness analysis, wide-table construction,
tests and assertions, debugging Pixi and Quarto, explaining statistical and
data concepts, and identifying edge cases to review.

The assistance reduced repetitive coding work, suggested useful checks, exposed
edge cases, and helped explain unfamiliar tooling. It also introduced risks:
AI can make unjustified clinical assumptions, and code can run successfully
while implementing the wrong data definition. The initial AI-assisted audit
missed some invalid pH values until a broader audit. Environment fixes also
became unnecessarily complicated during development. These experiences made
checking the definitions and outputs necessary.

The review process used reproducible tests, raw-record spot checks, and separate
calculations, including independent Python calculations for selected summaries.
Checks covered source-file counts and schemas, filename/internal RecordID
agreement, preservation of raw values, the distinction between hard-invalid
values and soft flags, boundaries at 24:00 and 48:00, long-to-wide spot checks,
and outcome joins. Figures were visually inspected. Locked Pixi execution and
a final fresh-clone reproduction were used to check reproducibility and compare
the long and wide CSV hashes with the accepted baseline.

Many checks were themselves developed or run with AI assistance; this is not
independent human validation of every generated line. Automated checks and
separate calculations can expose mistakes, but do not replace my responsibility
to personally read the final report and confirm that this disclosure accurately
reflects my work before submitting.

## Session information

```{r}
#| label: session-information
sessioninfo::session_info()
```
